---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# chronicler

<!-- badges: start -->
<!-- badges: end -->

Easily add logs to your functions.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("b-rodrigues/chronicler")
```

## Introduction

{chronicler} allows you to decorate functions to make them provide enhanced output:


```{r example}
library(chronicler)

r_sqrt <- record(sqrt)

a <- r_sqrt(1:5)

```

Object `a` is now an object of class `chronicle`. The value of the `sqrt()` function applied to its arguments
can be obtained using `pick()`:

```{r}
pick(a, "value")
```

A log also gets generated and can be read using `read_log()`:

```{r}
read_log(a)
```
This is especially useful for objects that get created using multiple calls:

```{r}

r_sqrt <- record(sqrt)
r_exp <- record(exp)
r_mean <- record(mean)

b <- 1:10 |>
  r_sqrt() |>
  bind_record(r_exp) |>
  bind_record(r_mean)

```

```{r}
read_log(b)

pick(b, "value")
```

## Composing decorated functions

`bind_record()` is used to pass the output from one decorated function to the next.

`record()` works with any function:

```{r}
library(dplyr)

r_group_by <- record(group_by)
r_select <- record(select)
r_summarise <- record(summarise)
r_filter <- record(filter)

output <- starwars %>%
  r_select(height, mass, species, sex) %>%
  bind_record(r_group_by, species, sex) %>%
  bind_record(r_filter, sex != "male") %>%
  bind_record(r_summarise,
              mass = mean(mass, na.rm = TRUE)
              )

```

```{r}
read_log(output)
```

The value can then be accessed and worked on as usual using `pick()`:

```{r}
pick(output, "value")
```

This package also ships with a dedicated pipe, `%>=%` which you can use instead of `bind_record()`:

```{r}
output_pipe <- starwars %>%
  as_chronicle() %>=%
  r_select(height, mass, species, sex) %>=%
  r_group_by(species, sex) %>=%
  r_filter(sex != "male") %>=%
  r_summarise(mass = mean(mass, na.rm = TRUE))

```

```{r}
pick(output_pipe, "value")
```

Objects of class `chronicle` have their own print method:

```{r}
output_pipe
```

## Condition handling

By default, errors and warnings get caught and composed in the log:

```{r}

errord_output <- starwars %>%
  as_chronicle() %>=%
  r_select(height, mass, species, sex) %>=% 
  r_group_by(species, sx) %>=% # typo, "sx" instead of "sex"
  r_filter(sex != "male") %>=%
  r_summarise(mass = mean(mass, na.rm = TRUE))

```

```{r}
errord_output
```

Reading the log tells you which function failed, and with which error message:

```{r}
read_log(errord_output)
```

It is also possible to only capture errors, or catpure errors, warnings and messages using
the `strict` parameter of `record()`

```{r}
# Only errors:

r_sqrt <- record(sqrt, strict = 1)

r_sqrt(-10) |>
  read_log()

# Errors and warnings:

r_sqrt <- record(sqrt, strict = 2)

r_sqrt(-10) |>
  read_log()

# Errors, warnings and messages

my_f <- function(x){
  message("this is a message")
  10
}

record(my_f, strict = 3)(10) |>
                         read_log()

```

## Advanced logging

You can provide a function to `record()`, which will be evaluated on the output. This makes it possible
to, for example, monitor the size of a data frame throughout the pipeline:

```{r}
r_group_by <- record(group_by)
r_select <- record(select, .g = dim)
r_summarise <- record(summarise, .g = dim)
r_filter <- record(filter, .g = dim)

output_pipe <- starwars %>%
  as_chronicle() %>=%
  r_select(height, mass, species, sex) %>=%
  r_group_by(species, sex) %>=%
  r_filter(sex != "male") %>=%
  r_summarise(mass = mean(mass, na.rm = TRUE))

```

The `$log_df` element of a `chronicle` object contains detailled information:

```{r}
pick(output_pipe, "log_df")
```

It is thus possible to take a look at the output of the function provided (`dim()`):

```{r}
output_pipe$log_df[, c("function", "g")]
```

We can see that the dimension of the dataframe was (87, 4) after the call to `select()`, (23, 4)
after the call to `filter()` and finally (9, 3) after the call to `summarise()`.

## Thanks

I’d like to thank [armcn](https://github.com/armcn), [Kupac](https://github.com/Kupac) for their
blog posts ([here](https://kupac.gitlab.io/biofunctor/2019/05/25/maybe-monad-in-r/)) and 
packages ([maybe](https://armcn.github.io/maybe/)) which inspired me to build this package.
