---
title: "Development actions history"
output: html_document
editor_options: 
  chunk_output_type: console
---

All commands that you use to use when developing packages...

# First time just after creating the project

- Fill the following chunk to create the DESCRIPTION of your package

```{r description}
# Describe your package
fusen::fill_description(
  pkg = here::here(),
  fields = list(
    Title = "Add Logging to Functions",

    Description = "Decorate functions to make them return enhanced output. The
  enhanced output consists in an object of type 'chronicle' containing the
  result of the function applied to its arguments, as well as a log detailing
  when the function was run, what were its inputs, what were the errors (if the
  function failed to run) and other useful information. Tools to handle
  decorated functions are included, such as a forward pipe operator that makes
  chaining decorated functions possible.",

    Version = "0.2.1.9999",

    `Authors@R` = c(
    person(given = "Bruno", family = "Rodrigues", , "bruno@brodrigues.co", role = c("aut", "cre", "cph"),
           comment = c(ORCID = "0000-0002-3211-3689")),
    person(given = "Matouš", family = "Eibich", "matous.eibich@seznam.cz", role = "ctb")
     )
    )
  )
)
# Define License with use_*_license()
usethis::use_gpl3_license()
```

# All-in-one function to deploy publicly on GitHub

Either use this GitHub all-in-one function or choose the steps one-by-one in the following sections.
See `vignette("share-on-a-github-website", package = "fusen")` for more information.

```{r, eval=FALSE}
# _All-in-one share of your package and its documentation on GitHub
fusen::init_share_on_github()
```


# Start using git

```{r, eval=FALSE}
usethis::use_git()
# Deal with classical files to ignore
usethis::git_vaccinate()
# Use main for primary branch
usethis::git_default_branch_rename()
```

# Set extra sources of documentation

```{r, eval=FALSE}
# Install a first time
remotes::install_local()
# README
usethis::use_readme_rmd()
# NEWS
usethis::use_news_md()
```

Add codemeta file:

```{r, eval = F}
codemetar::write_codemeta()

```

**From now, you will need to "inflate" your package at least once to be able to use the following commands. Let's go to your flat template, and come back here later if/when needed.**


# Package development tools
## Use once

```{r, eval=FALSE}
# Pipe
#usethis::use_pipe()

# package-level documentation
usethis::use_package_doc()

# GitHub
# Add your credentials for GitHub
usethis::create_github_token()
gitcreds::gitcreds_set()

# _Or one step at a time
# Send your project to a new GitHub project (included in `init_share_on_github()`)
usethis::use_github()

# Set Continuous Integration
# _GitHub (included in `init_share_on_github()`)
usethis::use_github_action_check_standard()
usethis::use_github_action("pkgdown")
usethis::use_github_action("test-coverage")
# _GitLab
gitlabr::use_gitlab_ci(type = "check-coverage-pkgdown")
```

```{r development-inflate, eval=FALSE}
# Run but keep eval=FALSE to avoid infinite loop
# Execute in the console directly
fusen::inflate(flat_file = "dev/z-save_avia.Rmd",
               vignette_name = "Developer Vignette: Including the avia data to the package",
               overwrite = TRUE)
```

```{r}
local({
  knitr::knit("vignettes/z-developers-vignette-including-the-avia-data-to-the-package.Rmd",
              output = tempfile())
  usethis::use_data(avia,
                    overwrite = TRUE)
})

```

```{r}
fusen::inflate(flat_file = "dev/flat_data_doc.Rmd",
               vignette_name = NA,
               overwrite = TRUE)
```

```{r}
fusen::inflate(flat_file = "dev/flat_chronicle.Rmd",
               vignette_name = NA,
               overwrite = TRUE)
```

```{r}
fusen::inflate(flat_file = "dev/flat_ggplot.Rmd",
               vignette_name = NA,
               overwrite = TRUE)
```

```{r}
fusen::inflate(flat_file = "dev/flat_zzz.Rmd",
               vignette_name = NA,
               overwrite = TRUE)
```

# Vignettes for users

```{r}
fusen::inflate(flat_file = "dev/maybe-monad.Rmd",
               vignette_name = "The Maybe monad",
               overwrite = TRUE)
```

```{r}
fusen::inflate(flat_file = "dev/advanced-topics",
               vignette_name = "A non-mathematician's introduction to monads",
               overwrite = TRUE)
```

```{r}
fusen::inflate(flat_file = "dev/read-world-example",
               vignette_name = "A real world example",
               overwrite = TRUE)
```



## Use everytime needed

```{r, eval=FALSE}
# Simulate package installation
pkgload::load_all()

# Generate documentation and deal with dependencies
attachment::att_amend_desc()

# Check the package
devtools::check()

```

# Share the package

```{r, eval=FALSE}
# set and try pkgdown documentation website locally
usethis::use_pkgdown()
pkgdown::build_site()

# build the tar.gz with vignettes to share with others
devtools::build(vignettes = TRUE)

# Share your package and its documentation on GitHub
# usethis::create_github_token()
# gitcreds::gitcreds_set()
fusen::init_share_on_github()
```

