---
title: "A non-mathematician's introduction to monads"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A non-mathematician's introduction to monads}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(chronicler)
library(testthat)
```

# Introduction

`{chronicler}` is an implementation of a logger monad. A monad is a computation device that offers
two things:

- the possibility to decorate functions so they can provide additional output without having to touch the function's core implementation;
- a way to compose these decorated functions;

To understand what a monad is, I believe it is useful to explain what sort of problem monads solve.

Suppose for instance that you wish for your functions to provide a log when they're run. If your
function looks like this:

```{r}
my_sqrt <- function(x){

  sqrt(x)

}

```

Then you would need to rewrite it like this:

```{r}
my_sqrt <- function(x, log = ""){

  list(sqrt(x),
       c(log,
         paste0("Running sqrt with input ", x)))

}

```

There are two problems with such an implementation:

- we need to rewrite every function;
- these functions don't compose.

What do I mean with "these functions don't compose"? Consider another such function `my_log()`:

```{r}
my_log <- function(x, log = ""){

  list(log(x),
       c(log,
         paste0("Running log with input ", x)))

}

```

`sqrt()` and `log()` compose, or rather, they can be chained:

```{r}
10 |>
  sqrt() |>
  log()

```

while this is not true for `my_sqrt()` and `my_log()`:

```{r, eval = FALSE}
10 |>
  my_sqrt() |>
  my_log()

```

```
Error in log(x) (from #3) : non-numeric argument to mathematical function
```

This is because `my_log()` expects a number, not a list which is what `my_sqrt()` returns.

A "monad" is what we need to solve these two problems. The first problem, not having to rewrite
every function, can be tackled using [function factories](https://adv-r.hadley.nz/function-factories.html).
Let's write one for our problem:

```{r}
log_it <- function(.f, ..., log = NULL){

  fstring <- deparse(substitute(.f))

  function(..., .log = log){

    list(result = .f(...),
         log = c(.log,
                 paste0("Running ", fstring, " with argument ", ...)))
  }
}

```

We can now create our functions easily:

```{r}
l_sqrt <- log_it(sqrt)

l_sqrt(10)

l_log <- log_it(log)

l_log(10)

```

We can call `l_sqrt()` and `l_log()` *decorated* functions and the values they return *monadic* values.

The second issue remains though; `l_sqrt()` and `l_log()` can't be composed/chained. To solve
this issue, we need another function, called `bind()`:

```{r}
bind <- function(.l, .f, ...){

  .f(.l$result, ..., .log = .l$log)

}
```

It is now possible to compose `l_sqrt()` and `l_log()`:

```{r}
10 |>
  l_sqrt() |>
  bind(l_log)

```

We can check that the result is correct by running the `sqrt()` and `log()`:

```{r}
log(sqrt(10))
```

This solution of using a function factory and defining a helper function to make the decorated
functions compose is what constitutes a monad, but strictly speaking, this is not precisely correct.
It can be interesting to see the actual definition from the programming language Haskell, which
is a pure functional programming language where monads *must* be used to solve certain issues:

*Monads can be viewed as a standard programming interface to various data or control structures, which is captured by Haskell's <code>Monad</code> class. All the common monads are members of it:*
<div class="mw-highlight mw-content-ltr" dir="ltr"><pre><span class="kr">class</span> <span class="kt">Monad</span> <span class="n">m</span> <span class="kr">where</span>
  <span class="p">(</span><span class="o">&gt;&gt;=</span><span class="p">)</span>  <span class="ow">::</span> <span class="n">m</span> <span class="n">a</span> <span class="ow">-&gt;</span> <span class="p">(</span>  <span class="n">a</span> <span class="ow">-&gt;</span> <span class="n">m</span> <span class="n">b</span><span class="p">)</span> <span class="ow">-&gt;</span> <span class="n">m</span> <span class="n">b</span>
  <span class="p">(</span><span class="o">&gt;&gt;</span><span class="p">)</span>   <span class="ow">::</span> <span class="n">m</span> <span class="n">a</span> <span class="ow">-&gt;</span>  <span class="n">m</span> <span class="n">b</span>         <span class="ow">-&gt;</span> <span class="n">m</span> <span class="n">b</span>
  <span class="n">return</span> <span class="ow">::</span>   <span class="n">a</span>                 <span class="ow">-&gt;</span> <span class="n">m</span> <span class="n">a</span>
</pre></div>

(Source: [Monad](https://wiki.haskell.org/Monad))

This definition is quite cryptic, especially if you don't know Haskell, but what this means is that
a `Monad` (in Haskell) comes with three methods:

- `>>=` which is what we called `bind()`;
- `>>` which I didn't bother implementing, because it's not really needed for understanding what a monad is;
- and `return`. Don't be confused by the name, this has nothing to do with the `return()` we use inside functions to return a value. `return` is a function that coerces a value into a monadic value.

While we didn't implement `return` (also called `unit`, which is also not a good name), our 
function factory `log_it()` does `return`/`unit`'s job as it *returns* `m f(a)` instead of `m a` 
(where `m` is our monad), so that is good enough. But I will implement `unit()` below.

Finally, you might read sometimes that monads are objects that have a `flatmap()` method. I think
that this as well is not strictly correct and very likely an oversimplification. But what is 
`flatmap()` anyways? In practical terms, it is equivalent to `bind()`, but it is how you get there
that's different. To implement `flatmap()` two additional functions are needed: `fmap()` and 
`flatten()` (which is quite often called `join()`, but this has nothing to do with *joining*
data frames, so I used `flatten()` instead).

`fmap()` is a function that takes a monadic value as an argument and an undecorated function and
applies this undecorated function to the monadic value:

```{r}
fmap <- function(m, f, ...){

  fstring <- deparse(substitute(f))

  list(result = f(m$result, ...),
       log = c(m$log,
               paste0("fmapping ", fstring, " with arguments ", paste0(m$result, ..., collapse = ","))))
}

```

Let's first define a monadic value:

```{r}
# This is `unit()`'s implementation:
unit <- log_it(identity)

(m <- unit(10))
```

Let's now use `fmap()` to apply a non-decorated function to `m`:

```{r}
fmap(m, log)
```

Great, now what about `flatten()` (or `join()`)? Suppose that instead of `log()` we used `l_log()`
with `fmap()`:

```{r}
fmap(m, l_log)
```

As you can see from the output, this produced a nested list. We would like `flatten()/join()` to take care 
of this for us. So this could be an implementation of `flatten()`:

```{r}
flatten <- function(m){

  list(result = m$result$result,
       log = c(m$log))

}

```

Let's try now:

```{r}
flatten(fmap(m, l_log))
```

Great! Now, as explained earlier, `flatmap()` and `bind()` are the same thing. But we have implemented
`flatten()` and `fmap()`, how does these two functions related to `flatmap()`? It turns out that
`flatmap()` is the composition of `flatten()` and `fmap()`:

```{r}
# I first define a composition operator
composition <- function(f, g){

  function(...){

    f(g(...))

  }

}

# I now compose flatten() an fmap()
flatmap <- composition(flatten, fmap)

```

So this means that we can now replace:

```{r}
10 |>
  l_sqrt() |>
  bind(l_log)

```

by:

```{r}
10 |>
  l_sqrt() |>
  flatmap(l_log)

```

and we get the same result (well, not quite, since the log is different). I prefer explaining 
monads using `bind()`, because `bind()` comes as a natural solution to the problem of decorated
functions not composing. Not so with `flatmap()`.

Before continuing with the final part of this introduction, I just want to share with you that
lists are also monads. We have everything we need: `as.list()` is `unit()`, `purrr::map()` is `fmap()`
and `purrr::flatten()` is `flatten()`. This means we can obtain `flatmap()` from composing
`purrr::flatten()` and `purrr::map()`:

```{r}
# Since I'm using `{purrr}`, might as well use purrr::compose() instead of my own implementation
flatmap_list <- purrr::compose(purrr::flatten, purrr::map)

# Functions that return lists: they don't compose!
# no worries, we implemented `flatmap_list()`
list_sqrt <- \(x)(as.list(sqrt(x)))
list_log <- \(x)(as.list(log(x)))

10 |>
  list_sqrt() |>
  flatmap_list(list_log)

```

(thanks to [@armcn_](https://twitter.com/armcn_/status/1511705262935011330?s=20&t=UfwIjsqyOX7-UbTMBHOCuw)
for showing me this)


# Monadic laws

Monads need to satisfy the so-called "monadic laws". We're going to verify if the monad implemented
in `{chronicler}` satisfies these monadic laws.

## First law

The first law states that passing a monadic value to a monadic function using `bind()` 
(or in the case of the `{chronicler}` package `bind_record()`) or passing a value to a monadic
function is the same.

```{r}
a <- as_chronicle(10)
r_sqrt <- record(sqrt)
 
test_that("first monadic law", {
  expect_equal(bind_record(a, r_sqrt)$value, r_sqrt(10)$value)
})

```

Turns out that this is not quite the case here; the logs of the two objects will be slightly 
different. So I only check the value.

## Second law

The second law states that binding a monadic value to `return()` (called `as_chronicle()` in
this package, in other words, the function that coerces values to chronicler objects) does
nothing. Here again we have an issue with the log, that's why I focus on the value:

```{r}
test_that("second monadic law", {
  expect_equal(bind_record(a, as_chronicle)$value, a$value)
})

```

## Third law

The third law is about associativity; applying monadic functions successively or composing them
first gives the same result.

```{r}
a <- as_chronicle(10)

r_sqrt <- record(sqrt)
r_exp <- record(exp)
r_mean <- record(mean)

test_that("third monadic law", {
  expect_equal(
    (bind_record(a, r_sqrt)) |>
      bind_record(r_exp),
    a |>
    (\(x) bind_record(x, r_sqrt) |> bind_record(r_exp))(),
  tolerance = .01
  )
})

```

# flatmap() for `chronicle` objects 

For exhaustivity's sake, I check that I can get `flatmap_record()` by composing `flatten_record()` and
`fmap_record()`:

```{r}

r_sqrt <- record(sqrt)
r_exp <- record(exp)
r_mean <- record(mean)

a <- 1:10 |>
  r_sqrt() |>
  bind_record(r_exp) |>
  bind_record(r_mean)

flatmap_record <- purrr::compose(flatten_record, fmap_record)

b <- 1:10 |>
  r_sqrt() |>
  flatmap_record(r_exp) |>
  flatmap_record(r_mean)

identical(a$value, b$value)

```
