---
title: "Recording ggplot Objects with chronicler"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Recording ggplot Objects with chronicler}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(chronicler)
library(ggplot2)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

`{chronicler}` allows you to "record" ggplot objects so you can trace their
creation, and debug plots without immediately rendering them. In case of an
error, an "error plot" gets shown instead, which can be useful when rendering
documents. This way, plots that fail don't stop the whole rendering process.
Unlike typical functions, ggplots are lazily evaluated, so capturing their
construction requires special handling.

This vignette demonstrates how to record and inspect ggplot objects using
`record_ggplot()`.

# Basic Recording

Create a simple plot and record it:

```{r}
r_plot <- record_ggplot(
  ggplot(mtcars, aes(x = mpg, y = hp)) +
    geom_point() +
    geom_smooth(method = "lm")
)
```

The object `r_plot` is now a **chronicle**. You can access the actual ggplot object using `unveil()`:

```{r}
unveil(r_plot, "value")
```

And inspect the log to see what happened during construction:

```{r}
read_log(r_plot)
```

# Why record ggplots

The main benefit is to capture errors and return an "error" plot, which is a
simple plot that shows the error message that was logged during the build
process. This is especially useful when working on a document, because this way,
if the plot fails, rendering can continue. This allows you to preview documents
and fix plots later.

# Handling Errors and Warnings

Typos in aesthetic mappings or layer definitions will be captured in the log rather than stopping execution:

```{r}
r_bad <- record_ggplot(
  ggplot(mtcars, aes(x = mpg, y = hpp)) +  # typo: 'hpp'
    geom_point()
)

r_bad
read_log(r_bad)
```

# Tips

- Use `unveil(..., "value")` to extract the ggplot object for rendering.
- Use `read_log()` to review warnings, errors, and messages during plot construction.
- Stepwise recording is useful in complex plotting pipelines or when experimenting interactively.

# Conclusion

`record_ggplot()` provides a safe, traceable way to build ggplot objects, making
complex plots easier to debug and understand. By integrating ggplot recording
with `{chronicler}`'s logging and chaining capabilities, you can maintain
reproducible, debuggable plotting workflows.
